---
title: "Homework 8"
author: "Pascal Pilz, k12111234"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 19

Since we have 25 samples and we have one of fields is less than 5, we use Fisher's exact test.

```{r}
data <- matrix(c(4, 5, 13, 3), 2, 2)
fisher.test(data)
```

We can see that there is no significant difference.

\newpage

# Exercise 20

```{r}
data <- matrix(c(5, 1, 2, 6), 2, 2)
```

We want to compare the asymptotic Chi-square test, Chi-square with Yate's correction, and exact Fisher test.

Since we have a sample size of XX and cells with values of less than 5 we already know that the assumption of the Chi-square test are violated.

## Chi-square

```{r}
chisq.test(data, correct=FALSE)
```

## Chi-square with Yate's correction

```{r}
chisq.test(data, correct=TRUE)
```

## Fiher's exact test

```{r}
fisher.test(data)
```

## Conclusion

As we can, the p-value for the corrected Chi-square test and Fisher's exact test are quite close, whereas the uncorrected Chi-square test produces a drastically different result.

\newpage

# Exercise 21

```{r}
p <- 100
n <- 200

pre <- p / (p + n)

tp <- 90
fp <- 30
tn <- 170
fn <- 10

se <- tp / p
sp <- tn / n

ppv_1 <- tp / (tp + fp)
ppv_2 <- (se * pre) / (se * pre + (1 - sp) * (1 - pre))

npv_1 <- tn / (tn + fn)
npv_2 <- (sp * (1 - pre)) / (sp * (1 - pre) + (1 - se) * pre)

cat(sprintf(" sensitivity: %.3f\n", se),
    sprintf("specificity: %.3f\n", sp),
    "\n",
    sprintf("positive predictive value (excluding prevelance): %.3f\n", ppv_1),
    sprintf("positive predictive value (including prevelance): %.3f\n", ppv_2),
    "\n",
    sprintf("negative predictive value (excluding prevelance): %.3f\n", npv_1),
    sprintf("negative predictive value (including prevelance): %.3f\n", npv_2))
```



